\section*{Acknowledgments}
First of all, I would like to express my gratitude to my advisor
Prof. Paolo Frasconi for his expert guidance, for the important
insights he shared about the field of AI and the always new and
inspiring inputs about research.

Second, I would like to thank Leonardo Ventura, Prof. Gianni Amunni,
and \mbox{ISPRO} for giving me the opportunity to work with their data and
making this work of thesis possibile.

Thanks to Prof. S{\o}ren Brunak and Prof. Beatrice Lazzerini for
helping me advancing my 
research. 

I’m grateful to Prof. Enrico Vicario for introducing me to this
doctorate and to Prof. Simone Marinai for his precious advice.

A special thanks goes to the people who shared the AI-Team lab and the
Ph.D life with me. In alphabetical order: Daniele Baracchi, Samuele
Capobianco, Alessandro Lazzeri, Francesco Orsini, Giulia Pellegrini,
Dasara Shullani,
Alessandro Tibo, La Ode Toresano, Amin Zadenoori, and Zahra Ziran.

I would like to thank all my friends that were always there to support
me, root for me and cheer me up in moments of crisis and
doubt. Special mentions: Andrea Benassai, Vito Bonelli, Giulia
Bondielli, Giulia Bono, Tiberio Uricchio. Thanks to my dear family
Mauro, Rosa, and Simone for their presence and affection throughout my
whole life. Thanks to Fede for all the help that she gave me in this
thesis and in my whole life and for loving me dearly. Last but not
least, thanks to Spallina for accompanying me from the start to the
end of my academic path.

